---
title: "Generalized Linear Model Analysis of Animal Shelter Stay Duration"
author: "Group Number: 16"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
  revealjs:
    
    auto-stretch: false
  pdf: 
    fig-pos: "H"
    dev: pdf  
    fig-format: pdf
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| label: libraries
library(dplyr)
library(ggplot2)
library(readr)
```

---
title: "Analysis of Factors Influencing Animal Shelter Stay Duration"
author: "DAS Group 16"
date: "`r Sys.Date()`"
output: pdf_document
---

# Introduction

This analysis focuses on understanding the factors that influence the duration of an animal's stay at the **Dallas Animal Shelter** before reaching its final outcome. The dataset assigned to our group contains records of animals admitted to the shelter, including key characteristics and details about their stay.

The dataset consists of the following variables:

-   **Animal_type**: The type of animal admitted to the shelter (e.g., dog, cat).
-   **Month**: The month the animal was admitted (numerically coded, January = 1, February = 2, etc.).
-   **Year**: The year the animal was admitted.
-   **Intake_type**: The reason for the animal's admission (e.g., stray, owner surrender, confiscation).
-   **Outcome_type**: The final recorded outcome of the admitted animal (e.g., adoption, return to owner, euthanasia).
-   **Chip_Status**: Whether the animal had a microchip with owner information at the time of admission.
-   **Time_at_Shelter**: The number of days the animal spent in the shelter before a final outcome was determined.

The goal of this study is to analyze the impact of these factors on **Time_at_Shelter** using **Generalized Linear Models (GLM)**. By understanding which variables contribute to longer or shorter stays, we aim to provide insights that could help optimize shelter operations and improve animal welfare outcomes.

# Data Import

```{r}

# Read Data
animal <- read.csv("C:/Users/丁乐之2024/OneDrive/Documents/dataset16.csv")

# View the data overview
str(animal)
```

# Exploratory Data Analysis (EDA)

## Distribution of Time at Shelter

```{r time-shelter-distribution}
# Histogram of time_at_shelter
animal %>%
  ggplot(aes(x = time_at_shelter)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Time Spent at the Shelter",
       x = "Days at Shelter",
       y = "Count") +
  theme_minimal()
```

· The distribution of Time_at_Shelter tends to be short-term, and most animals stay in shelters for a short time.

· Right-sided distribution, indicating that most animals will have final results quickly, but there are still some animals that stay longer.

· The KDE curve shows partial peaks, and there may be some key factors that affect residence time.

## Time at Shelter by Animal Type

```{r time-by-animal-type}
# Boxplot of time_at_shelter by Animal_type
animal %>%
  ggplot(aes(x = animal_type, y = time_at_shelter, fill = animal_type)) +
  geom_boxplot() +
  labs(title = "Time at Shelter by Animal Type",
       x = "Animal Type",
       y = "Days at Shelter") +
  theme_minimal() +
  theme(legend.position = "none")
```

· The residence time of different animal types (Animal_type) varies greatly:

· The median was significantly higher for certain animal types, indicating that they spent more time in shelters.

· Possible explanations: Certain animal types are more likely to be adopted or find owners more easily.

## Time at Shelter by Intake Type

```{r time-by-intake-type}
# Boxplot of Time_at_Shelter by Intake_type
animal %>%
  ggplot(aes(x = intake_type, y = time_at_shelter, fill = intake_type)) +
  geom_boxplot() +
  labs(title = "Time at Shelter by Intake Type",
       x = "Intake Type",
       y = "Days at Shelter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

· Reason of entry (Intake_type) has significant influence on Time_at_Shelter:

· Animals with certain entry reasons stay longer, such as abandonment or medical reasons.

· Animals that come in for other reasons may find their way faster, such as animals that come in because their owners are looking for them may leave faster.

## Time at Shelter by Outcome Type

```{r time-by-outcome-type}
# Boxplot of Time_at_Shelter by Outcome_type
animal %>%
  ggplot(aes(x = outcome_type, y = time_at_shelter, fill = outcome_type)) +
  geom_boxplot() +
  labs(title = "Time at Shelter by Outcome Type",
       x = "Outcome Type",
       y = "Days at Shelter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

· The distribution of Time_at_Shelter tends to be short-term, and most animals stay in shelters for a short time.

· Right-sided distribution, indicating that most animals will have final results quickly, but there are still some animals that stay longer.

· The KDE curve shows partial peaks, and there may be some key factors that affect residence time.

## Time at Shelter by Chip Status

```{r time-by-chip-status}
# Boxplot of Time_at_Shelter by Chip_Status
animal %>%
  ggplot(aes(x = chip_status, y = time_at_shelter, fill = chip_status)) +
  geom_boxplot() +
  labs(title = "Time at Shelter by Microchip Status",
       x = "Microchip Status",
       y = "Days at Shelter") +
  theme_minimal()
```

· Chip_Status has an effect on Time_at_Shelter:

· Animals with chips tend to stay shorter, possibly because the chip information helps find the owner faster.

· Animals without chips may be more dependent on shelter arrangements, resulting in longer stays.

## Time at Shelter by Year

```{r time-by-year}
# Boxplot of Time_at_Shelter by Year
animal %>%
  ggplot(aes(x = as.factor(year), y = time_at_shelter, fill = as.factor(year))) +
  geom_boxplot() +
  labs(title = "Time at Shelter by Year",
       x = "Year",
       y = "Days at Shelter") +
  theme_minimal() +
  theme(legend.position = "none")
```

· Year affects Time_at_Shelter:

· The median varies from year to year and may be affected by policy or how shelters are managed.

· In recent years, there may have been more efficient ways to relocate, so that the length of stay has fallen.

## Time at Shelter by Month

```{r time-by-month}
# Boxplot of Time_at_Shelter by Month
animal %>%
  ggplot(aes(x = as.factor(month), y = time_at_shelter, fill = as.factor(month))) +
  geom_boxplot() +
  labs(title = "Time at Shelter by Month",
       x = "Month",
       y = "Days at Shelter") +
  theme_minimal()
```

· Month (Month) has an effect on Time_at_Shelter:

· Longer stays may occur in certain months (such as summer and winter), which may be related to seasonal adoption trends or changes in stray animal populations.

· Further analysis can be combined with weather, holidays and other factors to explore deeper causes.

## Here we preform GLM analysis

```{r}
library(tidyverse)
library(moderndive)
library(gapminder)
library(sjPlot)
library(stats)
library(jtools)
library(janitor)

```

# data summary description

```{r}

animal$animal_type<-as.factor(animal$animal_type)
animal$month<-as.factor(animal$month)
animal$year<-as.factor(animal$year)
animal$intake_type<-as.factor(animal$intake_type)
animal$outcome_type<-as.factor(animal$outcome_type)
animal$chip_status<-as.factor(animal$chip_status)
animal$time_at_shelter<-as.factor(animal$time_at_shelter)

animal %>%
  tabyl(intake_type,time_at_shelter) %>%
  adorn_percentages() %>%
  adorn_pct_formatting() %>%
  adorn_ns() # To show original counts

ggplot(data = animal, aes(x = time_at_shelter, group = intake_type)) +
 geom_bar(aes(y = ..prop.., fill = time_at_shelter),
 stat = "count", position = "dodge") +
 labs(x = "number of days an animal spends in the shelter ", y = "intake_type")

```

# build model and fitted data

```{r}
 model.time_at_shelter <- glm(time_at_shelter ~ animal_type + month + year + intake_type+ outcome_type+chip_status, data = animal,
 family = binomial(link = "logit"))
 model.time_at_shelter %>%
 summary()
```

# Odds of animals surviving over the number of day.

```{r}
 plot_model(model.time_at_shelter, show.values = TRUE,
 title = "", show.p = FALSE, value.offset = 0.25)


 
```

# probability of number of days an animal spends in the shelter.

```{r}
model.time_at_shelter.coef.logodds <- model.time_at_shelter %>%
                            summary() %>%
                            coef()
plogis(model.time_at_shelter.coef.logodds["(Intercept)", "Estimate"])


```



